### 2018-12-05 submission/manuscript.Rmd
Remove 15 cycle data

16S rRNA gene sequencing is a widely used tool for surveying the structure of microbial communities
* Numerous sources of bias ranging from sample preservation, extraction, primers, PCR, and sequencing
* PCR and sequencing errors can also influence our ability to accurately describe the structure of the communities
* Within a study, it is unlikely that samples would be differentially affected by these sources of bias; however, if we want to compare the results of one study to another, the results are likely to be confounded by the differences in these biases and sources of error
* There has been great interest in developing bioinformatics tools to account for errors and limited interest in accounting for bias
* There has also been considerable interest in considering how laboratory methods influence biases and errors.
* One area that has not received extensive attention has been the influence of the polymerase on bias and error.

Factors known to influence PCR bias
* Fragment length
* GC content
* Flanking DNA sequence
* Polymerase?
* Complex communities?
* Number of cycles

Factors known to influence chimera rate
* Fragmentation of DNA
* DNA concentration
* Similarity of potential parents
* PCR extension time
* Number of cycles

Factors known to influence polymerase error rates
* Number of cycles
* Proof-reading
* Enzymology

Early work focused on simple mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation.
* Have a poor understanding of how polymerase impacts these problems
* Have a poor understanding of how these problems manifest in more complex communities

The bacterial community is reported to vary between case and control for a number of diseases [@Turnbaugh2008; @Baxter2016; @Sze2015; @Bonfili2017]. However, for diseases like obesity, the taxa identified have varied widely depending on the study [@Turnbaugh2008; @Zupancic2012]. Although some of the differences could be attributed to study design, error introduced during the 16S rRNA gene sequencing workflow could also be another source of variation. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. The preservation and extraction stages of the 16S rRNA gene sequencing workflow have been the most extensively studied [@contamination_Salter2014; @preservation_Song_2016; @storage_Bassis_2017; @review_Kim_2017]. For preservation and extraction stages of the workflow, it has been consistently found that there are errors in the observed bacterial community composition based on the kits used, but that these differences are smaller than the overall biological difference measured between samples with different kits [@preservation_Song_2016; @storage_Bassis_2017]. Since these studies use the same PCR approach while varying preservation or extraction method, the contribution of PCR error to this overall workflow is not as well characterized.

There is a large body of literature that shows there are errors due to primer and number of cycles chosen for the PCR stage of 16S rRNA gene sequencing [@Eckert1991; @Burkardt2000]. Primers have variable region dependent binding affinities which causes an inability to detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Another source of error is the selective amplification of AT-rich over GC-rich sequences which exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC content [@polz_bias_1998]. Many of these sources of error are made worse as the number of cycles increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, both amplification error and non-specific amplification (e.g. incorrect amplicon size products) also can increase as the number of cycles increases. This will increase the number of Operational Taxonomic Units (OTUs) observed and drastically change the values obtained from commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, as the number of cycles increases more chimeras can form from an aborted extension step that causes a priming error and subsequent secondary extension [@Haas2011]. These chimeras will artificially increase community diversity by increasing the number of OTUs that are observed [@Haas2011]. In addition to these sources of errors, there also are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. For simplicity, the use of the term 'error' within the context of this study will be used as a broad term for changes in the downstream sequencing results due to the amplification step (e.g. higher GC content genomes, base substitutions, chimeras, changes from the expected relative abundance, etc.).

Interestingly, the influence different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other sources of PCR-based error. A recent study found differences in the number of OTUS and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet, within this study there was no comparison made between different high fidelity DNA polymerases on the same samples. The authors instead focused their efforts on testing specific types of amplification methods. Additionally, the fecal samples that were analyzed, as a model of a more complex community, may overestimate the biological variation in human samples due to the comparison of captive versus semi-captive red-shanked doucs. Due to these gaps, it makes sense to extend the previous line of inquiry to include a more detailed examination of two specific areas. First, whether differences in PCR error are detected based on number of cycles used and high fidelity polymerases. Second, whether these differences in error are enough to obscure smaller biological signals than what has been previously reported [@taq_Gohl_2016].

This study will investigate whether there are error differences that are dependent on high fidelity DNA polymerases and whether these differences obscure the biological variation in the fecal bacterial communities between individuals. We accomplish this by first examining GC-based amplification error, sequence error (e.g. substitutions), chimera prevalence, the number of OTUs, Bray-Curtis index, and OTU-based differences at varying number of cycles in five different high fidelity DNA polymerases in mock communities. In addition to these metrics we also created Random Forest models of the mock community to assess how successful classification based on 16S rRNA gene sequencing of number of cycles and type of polymerase could be. We then assessed if similar differences in the number of OTUs, Bray-Curtis index, and OTUs could be detected in human fecal samples. Additionally, we also created Random Forest models to assess how successful they could be at classifying fecal samples by number of cycles, polymerase used, or individual. Collectively, our observations suggest that although detectable differences occur based on number of cycles or high fidelity DNA polymerase used, they are smaller than the biological differences between individuals.

## pat's final paragraph
We addressed X potential areas where the type of PCR polymerase was hypothesized to impact the results of a microbiome analysis. These included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process. To characterize these factors we sequenced a mock community with known sequences and community structure and human fecal samples with unknown sequences and community structures.

Three factors can impact whether an organism's 16S rRNA gene sequence is correctly represented. These include mis-incorporated bases during the PCR amplification phase, mis-incorporated bases during sequencing, and creation of chimeras during the PCR amplification phase.

We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform. This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol. The mock community is commercially available and was made up of genomic DNA from 8 bacteria. The benefit of using this mock community is that the sequences of the 16S rRNA genes from these bacteria is known. This allowed us to confidently identify any chimeras and to calculate the sequencing error rate.

We selected five high fidelity DNA polymerases to amplify the V4 region of the 16S rRNA gene from a mock community containing genomic DNA from 8 previously sequenced bacteria. The genomic DNA from the mock community was amplified using 15, 20, 25, 30, and 35 rounds of amplification using one of five different polymerases. Each condition was replicated four times and reads for each condition were pooled. Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur. Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses. The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur. This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence. The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.

@@

***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels appeared to be small relative to the overall differences between samples. To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification. We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor. Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions. We also performed a principle coordinates analysis of the Bray-Curtis distances showed distinct clusters by donor. Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR. Both of these results indicate that for these samples, the differences were larger than the difference between conditions. These results suggest that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important.

Generate PCoA diagrams
Alpha diversity figures

@@

@@

Based on polymerase used, our observations generally found no difference when using the Bray-Curtis index to measure the bacterial community. Specifically, there was no difference in distance between successive 5-cycle increments within samples between the polymerases when using the Bray-Curtis index [Figure 4]. One possible reason for this outcome is that many of the OTUs generated by polymerase dependent error rates and chimera prevalence are likely not highly abundant, allowing the Bray-Curtis index to be able to successfully down-weight these OTUs [@bc_index_Minchin1987]. The choice of downstream diversity metric could be an important consideration in helping to mitigate some of the observed polymerase dependent differences in error rate and chimera prevalence. Metrics that solely use presence/absence of OTUs (e.g. Jaccard [@Real1996], richness) may be less robust to polymerase dependent error rates and chimera prevalence. When choosing a distance metric, careful consideration of the errors introduced from the PCR step of the 16S rRNA gene sequencing workflow need to be taken into account.

Collectively, our data shows that using distinct high fidelity DNA polymerases will result in a different number of OTUs being detected. The differences in the number of OTUs are primarily due to chimera prevalence and sequence error rates that are distinct to the specific polymerase. Despite this variation in the number of OTUs, our Random Forest models were able to correctly classify individuals regardless of the number of cycles and polymerase used. These results suggest that even though there is a large variation in the high fidelity DNA polymerases used across studies, biological variation similar in effect to fecal bacterial community composition differences between individuals will still be consistent.

@@

>>>>

16S rRNA gene sequencing is a powerful and widely used tool for surveying the structure of microbial communities. This approach has exploded in popularity with the advent of high throughput sequencing where it is possible to characterize numerous samples with thousands of sequences per sample. Numerous factors can impact how a natural community is represented by the sequencing data including the method of acquiring samples, storage conditions, extraction methods, amplification conditions, sequencing method, and analytical pipeline. The increased sampling depth that is now available relative to previous Sanger-based methods is expected to compound the impacts of an investigator's choices and the interpretation of their results.

One step in the generation of 16S rRNA gene sequence data that has been long known to have a significant impact on the description of microbial communities is the choice of conditions for PCR amplification. Factors such as the choice of primers have an obvious impact on which populations will be amplified. However, a variety of PCR artifacts can also impact the perception of a community including the formation of chimeras, misincorporation of nucleotides, preferential amplification of populations leading to bias, and accumulation of random amplification events leading to PCR drift. Many bioinformatic tools have been developed to identify chimeras, there are significant sensitivity and specificity tradeoffs. Others have attempted to account for PCR bias using modeling approaches; however, these have been developed for idealized situations. Laboratory-based solutions to minimize chimera formation have also been proposed such as minimizing the amount of template DNA in the PCR, minimizing the number of rounds of PCR, minimizing the amount of shearing in the template DNA, and using DNA polymerases that have a proof-reading ability. To minimize PCR drift, some investigators will pool technical replicate PCRs hoping to average out the drift. Other factors that have been shown to impact the formation of PCR artifacts are outside the control of an investigator including the fraction of DNA bases that are guanines or cytosines, the variation in the length of the targeted region across the community, the sequence of the DNA that flanks the template, and the genetic diversity of the community. Early investigations of the factors that lead to the formation of PCR artifacts focused on analyzing binary mixtures of genomic DNA and 16S rRNA gene fragments to explore PCR biases and chimera formation. Although these studies were instrumental in forcing researchers to be cautious about the interpretation of their results, we have a poor understanding of how these factors affect the formation of PCR artifacts in more complex communities.

The influence that the choice of DNA polymerase has on the formation of PCR artifacts has not been well studied. There has been a lot of recent interest in how the choice of the hypervariable region and data analysis pipelines impact the sequencing error rate; however, these studies use the same DNA polymerase in the PCR step and assume that the rate of misincorporation of nucleotides due to PCR are significantly smaller than those from the sequencing phase. There has been more limited interest in the impact that DNA polymerase choice has on the formation of chimeras. A recent study found differences in the number of OTUs and chimeras between normal and high fidelity DNA polymerases [@taq_Gohl_2016]. The authors of this study could reduce the difference between two polymerases by optimizing the annealing and extension steps within the PCR protocol [@taq_Gohl_2016]. Yet this optimization was specific for the community they were analyzing (i.e. captive and semi-captive red-shanked doucs) and assumes that if the two polymerases generate the same community structure that the community structure is correct. In fact, the community structure generated by both methods is not free of artifacts, but has the same artifacts. A challenge in these types of experiments is a priori knowledge of the true community representation. A mock community with known composition allows researchers to quantify the sequencing error rate, fraction of chimeras, and bias; however, mock communities have a limited phylogenetic diversity relative to natural communities. Natural communities, in contrast, have an unknown community composition making absolute measurements impossible. They can be used to validate results from mock communities and to understand the relative impacts of artifacts on the ability to differentiate biological and methodological sources of variation. Given the large number of DNA polymerases available to researchers it is unlikely that a specific recommendation is possible. Rather, the development of general best practices and understanding the impact of PCR artifacts on an analysis are desired.

This study investigated the impact of choice of high fidelity DNA polymerase and the number of rounds of amplification on the formation of PCR artifacts using a mock community and human stool samples. These PCR artifacts included (i) the effect of the polymerase on the error rate of the bases represented in the final sequences, (ii) the fraction of sequences that appeared to be chimeras and the ability to detect those chimeras, (iii) the bias of preferentially amplifying one fragment over another in a mixed pool of templates, and (iv) inter-sample variation in community structure of samples amplified with the same polymerase across the amplification process. To characterize these factors we sequenced a mock community of 8 organisms with known sequences and community structure and human fecal samples with unknown sequences and community structures. We sequenced the V4 region of the 16S rRNA genes from a mock community by generating paired 250 nt reads on the Illumina MiSeq platform. This region and sequencing approach was used because it has been shown to result in a relatively low sequencing error rate and is a widely used protocol.
To better understand the impact of DNA polymerase choice on PCR artifacts, we selected five high fidelity DNA polymerases and amplified the communities using 15, 20, 25, 30, and 35 rounds of amplification (Table 1). Collectively, our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals.

@@

***The choice of polymerase or the number of rounds of amplification have little impact on the relative interpretation of community-wide metrics of diversity.*** The biases that we observed at the population and community levels appeared to be small relative to the overall differences between samples. To study this further, we calculated alpha and beta-diversity metrics using the human stool samples for each of the polymerases and rounds of amplification. We calculated the number of observed OTUs, Shannon diversity, inverse Simpson diversity index for each condition and donor. Although there were clear differences between conditions, the relative ordering of the stool samples did not meaningfully vary across conditions. We also performed a principle coordinates analysis of the Bray-Curtis distances showed distinct clusters by donor. Within each cluster there were no obvious patterns related to the polymerase or number of rounds of PCR. However, when we performed the same analysis using the mock community, there were clear shifts in the represented community structures as the number of rounds of PCR increased. Using adonis, there were statistically significant differences in the community structures for the mock and stool community data (all P<0.001). Within the mock community data, the number of rounds of PCR explained the most variation (R^2^=0.XXX) followed by the choice of polymerase (R^2^=0.XXX). Similarly, within the stool community data, the subject explained the most variation (R^2^=0.XXX), followed by the number of rounds of PCR (R^2^=0.XXX) and the choice of polymerase (R^2^=0.XXX). Together, these results indicate that for a coarse analysis of communities, the choice of number of rounds of amplification or polymerase are not important, but that they must be consistent across samples.

@@

* Something like... our results suggest that the number of rounds and to a lesser degree the choice of DNA polymerase used in PCR impact the sequence data, the effects are consistent and are smaller than the biological differences between individuals

* be consistent within a study; do not interpret values like richness in absolute, but relative terms; take care in comparing results across studies

* GC, sequence length non-issues

* Other factors to investigate include impact of phylogenetic diversity, shearing, extension time, interactions of factors probably precludes ability to make definitive decisions.

* Interpret results in light of polymerase characteristics - Accuprime and Platinum have error rates that are ~10-times lower than that of Taq whereas Phusion, Q5, and Kappa are >100 times lower. These error rates track well with what we saw for sequencing error rates

* Minimize number of rounds of amplification - bias could be due to mis-determination of relative abundances by zymo

* Use multiple methods to triangulate on reality/validate claims

@@

@@

#Each condition was replicated four times and reads for each condition were pooled. Paired 250-nt sequencing reads were generated on an Illumina MiSeq sequencing platform and the reads were assembled using the make.contigs command in mothur. Any contigs that contained an ambiguous base call, mapped to the incorrect region of the 16S rRNA gene, or appeared to be a contaminant were removed from subsequent analyses. The sequencing error rates of the resulting contigs were quantified using the seq.error command in mothur. This command uses the true sequences from the mock community to generate all possible chimeras and removes any contigs that were at least three bases more similar to a chimera than to a reference sequence. The command then counts the number of substitutions, insertions, and deletions in the contig relative to the reference sequence.
